A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-09-28 rigid_bodies.tex
Avanços na soluçao analítica

\newenvironment{smallarray}[1]                          % small arrays
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{0.7}
	\arraycolsep=.13885em
	\hbox\bgroup$\array{@{}#1@{}}}
{\endarray$\egroup\egroup\,\null}

@@

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{smallarray}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{smallarray} \right].
\]

For every non-collinear rigid body (for conciseness, the only case considered here), there exists a particular frame of reference in which its inertia tensor is diagonal. The formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{smallarray}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{smallarray} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation. With no loss of generality, we can consider that $I_1 > I_2 > I_3$. Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g. Ref.~\onlinecite{Kopp2008}). Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric. The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).

@@

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Methods are available for computing the unit quaternion related to a given rotation matrix (e.g. Ref.~\onlinecite{Shepperd1978}). Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\[
\label{eq:A_from_q}
\mt A(\vt q) = \left[ \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray} \right].
\]

@@

Let us now introduce two new matrix-valued functions that depend on the quaternion components as
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \left[
\begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\right]
\quad \text{and} \quad
\mt C(\vt q) = \left[
\begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}
\right].
\end{equation}

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\[
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\]

@@

Hence, $\vt \pi$ is twice the quaternion-fixed version of the angular momentum ($\mt I \vt \omega$) and, as such, it is also orthogonal to $\vt q$. Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi}{8} + U(\vt r, \vt q).
\end{equation}

@@

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame. Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in quaternion-fixed frame of reference. Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:euler_eq_quat}
\end{align}
\end{subequations}

@@

Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eq.~\ref{eq:EDO_system}:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \vt F \\
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

@@

Before describing the method, we consider the transformati

@@

@@

 Once $\vt L$ is computed from the initial condition and the evolution of $\vt \omega$ is known (after solving the Euler equation), this fact constrains the possible evolution of $\vt q$. By premultiplying $\vt L$ by $\vt C$, one has that $\vt C \vt L = \vt B \mt I \vt \omega$. In other words, $\vt q$ must vary in such a way that the mapping of both $\vt L$ and $\mt I\vt \omega$ in its orthogonal plane always coincide. Then, resorting to Eq.~\ref{eq:product_B_vector} we can write
\begin{equation*}
\sum_{k=1}^3\left(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k\right){\vt q} = \vt 0,
\end{equation*}
where $\hat{\vt C}_k$ and $\hat{\vt B}_k$ are the permutation matrices defined in Appendix~\ref{sec:auxiliary_math}.
Therefore, $\vt q$ is constrained to lie in the null space of a matrix $\mt G = \sum_{k=1}^3(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k)$. The determinant of this matrix can be obtained as
\begin{equation*}
\det \mt G = (\tr{\vt \omega}{\mt I}^2{\vt \omega} - L^2)^2,
\end{equation*}
which is identically null in a free rotational motion. In fact, this is a requirement for a matrix to have a non-empty null space. Using a computer algebra package, it is straightforward to find out that $\vt G$ has a two-dimensional null space whose basis set is formed by the two vectors
\begin{equation*}
\vt v_1 = \left[\begin{smallarray}{c}
L_2 + I_2 \omega_2 \\
L_3 - I_3 \omega_3 \\
0 \\
L_1 - I_1 \omega_1
\end{smallarray}\right]
\quad \text{and} \quad
\vt v_1 = \left[\begin{smallarray}{c}
L_3 + I_3 \omega_3 \\
L_2 - I_2 \omega_2 \\
L_1 - I_1 \omega_1 \\
0
\end{smallarray}\right].
\end{equation*}

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\]
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

@@

It is worth noting that an analytical solution can also be derived for the system of equations $\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i$ for all $i \in [1,N]$ if all other variables are constant. These equations are equivalent in form to the one solved above, so that their solutions are
\begin{align*}
&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
&\eta_1(t) = \eta_1^0 + \alpha_1 t
\end{align*}

\begin{subequations}
\label{eq:solution_momenta}
\begin{align}
&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \quad \text{and} \label{eq:solution_p} \\
&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t. \label{eq:solution_pi}
\end{align}
\end{subequations}

@@

Recall that $\vt \pi$ is twice the quaternion-fixed version of the angular momentum vector. Thus, it is expected that Eq.~\ref{eq:euler_eq_quat} will result in the well-known Euler equation, which describes the evolution of the angular momentum in the body-fixed frame, if $\frac{1}{2} \tr{\mt B}$ is premultiplied to both sides. In order to show that this actually occurs, we start by carrying out the premultiplication to obtain
\[
\frac{1}{2} \tr{\mt B} \dot{\vt \pi} = \frac{1}{4} \tr{\mt B} \mt \Omega \vt \omega + \mt A \vt \tau.
\]

>>>>

\newenvironment{smallarray}[1]                          % small arrays
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{0.7}
	\arraycolsep=.13885em
	\hbox\bgroup$\left[\array{@{}#1@{}}}
{\endarray\right]$\egroup\egroup\,\null}

@@

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \begin{smallarray}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{smallarray}.
\]

For every non-collinear rigid body (for conciseness, the only case considered here), there exists a particular frame of reference in which its inertia tensor is diagonal. The formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \begin{smallarray}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{smallarray},
\]
where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation. With no loss of generality, we can consider that $I_1 > I_2 > I_3$. Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g. Ref.~\onlinecite{Kopp2008}). Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric. The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).

@@

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Methods are available for computing the unit quaternion related to a given rotation matrix (e.g. Ref.~\onlinecite{Shepperd1978}). Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\[
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray}.
\]

@@

Let us now introduce two new matrix-valued functions that depend on the quaternion components as
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations. We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\[
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\]

@@

Hence, $\vt \pi$ is twice the quaternion-fixed version of the angular momentum $\vt L = \mt I \vt \omega$ and, as such, it is also orthogonal to $\vt q$. Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi}{8} + U(\vt r, \vt q).
\end{equation}

@@

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame. Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in quaternion-fixed frame of reference. Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
\end{align}
\end{subequations}

@@

Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

@@

It is not possible Eq.~\ref{eq:EDO_system}

@@

Considering that $I_1 > I_2 > I_3$, the prefactors of the angular momenta $\omega_1$, $\omega_2$, and $\omega_3$ are, respectively,
\begin{align}
\mt A = \left[\begin{array}{ccc}
\frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}} & 0 & 0 \\
0 & \min(\lambda_1,\lambda_3) & 0 \\
0 & 0 & \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
\end{array}\right], \\
\end{align}
\begin{align}
A_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}}, \\
A_2 = \min(\lambda_1,\lambda_3), \; \text{and} \\
A_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}},
\end{align}
where
\begin{equation}
\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \;\; \text{and} \;\; \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
\end{equation}

The solution is
\begin{subequations}
\begin{align}
\omega_1(t) &= \begin{cases}
A_1\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
A_1\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
\end{cases} \\
\omega_2(t) &= A_2 \text{sn}(a t + b,k) \\
\omega_3(t) &= \begin{cases}
A_3\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
A_3\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
\end{cases}
\end{align}
\end{subequations}
where
\begin{align*}
&a = -\frac{(I_1-I_3)A_1A_3}{I_2 A_2},\\
&b = \text{sn}^{-1}\left(\frac{\omega_{2,0}}{A_2},k\right),\quad\text{and}\\
&k = \frac{A_2}{\text{max}(\lambda_1,\lambda_3)}
\end{align*}

@@

 Once $\vt L$ is computed from the initial condition and the evolution of $\vt \omega$ is known (after solving the Euler equation), this fact constrains the possible evolution of $\vt q$. By premultiplying $\vt L$ by $\vt C$, one has that $\vt C \vt L = \vt B \mt I \vt \omega$. In other words, $\vt q$ must vary in such a way that the mapping of both $\vt L$ and $\mt I\vt \omega$ in its orthogonal plane always coincide. Then, resorting to Eq.~\ref{eq:product_B_vector} we can write
\begin{equation*}
\sum_{k=1}^3\left(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k\right){\vt q} = \vt 0,
\end{equation*}
where $\hat{\vt C}_k$ and $\hat{\vt B}_k$ are the permutation matrices defined in Appendix~\ref{sec:auxiliary_math}.
Therefore, $\vt q$ is constrained to lie in the null space of a matrix $\mt G = \sum_{k=1}^3(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k)$. The determinant of this matrix can be obtained as
\begin{equation*}
\det \mt G = (\tr{\vt \omega}{\mt I}^2{\vt \omega} - L^2)^2,
\end{equation*}
which is identically null in a free rotational motion. In fact, this is a requirement for a matrix to have a non-empty null space. Using a computer algebra package, it is straightforward to find out that $\vt G$ has a two-dimensional null space whose basis set is formed by the two vectors
\begin{equation*}
\vt v_1 = \begin{smallarray}{c}
L_2 + I_2 \omega_2 \\
L_3 - I_3 \omega_3 \\
0 \\
L_1 - I_1 \omega_1
\end{smallarray}
\quad \text{and} \quad
\vt v_1 = \begin{smallarray}{c}
L_3 + I_3 \omega_3 \\
L_2 - I_2 \omega_2 \\
L_1 - I_1 \omega_1 \\
0
\end{smallarray}.
\end{equation*}

\begin{equation*}
\exp\left(-\begin{smallarray}{cc}
 \alpha(t) & \phi(t) \\
-\phi(t) & \alpha(t)
\end{smallarray}\right) = 
e^{-\alpha(t)} \begin{smallarray}{cc}
\cos \phi(t) & -\sin \phi(t) \\
\sin \phi(t) &  \cos \phi(t)
\end{smallarray}
\end{equation*}

Here,
\begin{equation}
\alpha(t) = \frac{1}{2}\int_0^t \frac{I_1 \omega_1^\prime(\tau)}{L - I_1\omega_1(\tau)}d\tau =
\frac{1}{2}\ln\left(\frac{L - I_1\omega_1}{L - I_1\omega_1^0}\right)
\end{equation}

\begin{equation}
\varphi(t) = \frac{1}{2}\int_0^t \frac{2 K_r + L \omega_1(\tau)}{L + I_1\omega_1(\tau)}d\tau
\end{equation}

After some algebra\cite{Celledoni_2008}
\begin{equation*}
\varphi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2 I_1 L}\int_0^t \frac{1}{1+(I_1/L)\omega_1(\tau)}d\tau.
\end{equation*}

If $\lambda_1 < \lambda_3$, then $\omega_1(t) = A_1 \text{cn}(\alpha t + \epsilon)$

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\begin{equation*}
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\end{equation*}
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

@@

It is worth noting that an analytical solution can also be derived for the system of equations $\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i$ for all $i \in [1,N]$ if all other variables are constant. These equations are equivalent in form to the one solved above, so that their solutions are
\begin{align*}
&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
&\eta_1(t) = \eta_1^0 + \alpha_1 t
\end{align*}

\begin{subequations}
\label{eq:solution_momenta}
\begin{align}
&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \quad \text{and} \label{eq:solution_p} \\
&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t. \label{eq:solution_pi}
\end{align}
\end{subequations}

@@

Recall that $\vt \pi$ is twice the quaternion-fixed version of the angular momentum vector. Thus, it is expected that Eq.~\ref{eq:EDO_pi} will result in the well-known Euler equation, which describes the evolution of the angular momentum in the body-fixed frame, if $\frac{1}{2} \tr{\mt B}$ is premultiplied to both sides. In order to show that this actually occurs, we start by carrying out the premultiplication to obtain
\[
\frac{1}{2} \tr{\mt B} \dot{\vt \pi} = \frac{1}{4} \tr{\mt B} \mt \Omega \vt \omega + \mt A \vt \tau.
\]
